Reproduce analyses reported in: *Phenotypic evolution is more constrained in simpler food webs*

### 2019-01-14 analyses/reproduce_analyses.Rmd
adding explanations for each code chunk and general tidying.

---
title: "Reproduce analyses reported in: *Phenotypic evolution is more constrained in simpler food webs*"
author: "Matthew A. Barbour"
date: "`r Sys.Date()`"
output: github_document
# Trying github_document for easy viewing on github
#  prettydoc::html_pretty:
#    theme: architect
#    highlight: github
fontsize: 12pt
---

@@

## Load required libraries ----
#library(evolqg) # need to load before dplyr
library(car)        # for homogeneity of variance test
library(tidyverse)  # for managing data
library(cowplot)    # pretty default ggplots
library(broom)      # for tidying multiple linear models
library(visreg)     # for quick visualizations of model effects
library(viridis)    # for color palette
library(lme4)       # for generalized linear mixed models
library(piecewiseSEM) # for calculating model R-square
library(latex2exp)  # using Latex notation for axes labels
#library(kableExtra)
library(stringr)
library(MVN)        # assessing multivariate normality assumption

@@

We write the model in a way the independently estimates the effect of food-web treatment, each trait, and all two-way and three-way statistical interactions, on larval survival.

@@

Note that the resulting estimates and confidence intervals are useful for determining whether trait-fitness relationships differ from zero, but not whether they differ between food-web treatments. For the later, we calculate the differences between each food-web treatment from the bootstrapped samples.

```{r Bootstrap food-web model, cache=TRUE, include=FALSE}
boot_foodweb_model <- bootMer(foodweb_model, FUN = fixef, nsim=n_boots_analysis, parallel="multicore", ncpus=32)
```

@@

```{r Bootstrap linear food-web model, cache=TRUE, include=FALSE}
boot_linear_foodweb_model <- bootMer(linear_foodweb_model, FUN = fixef, nsim=n_boots_analysis, parallel="multicore", ncpus=32)
```

To estimate biased selection on chamber diameter, we subset our data to only include multi-chambered galls where there was variability in larval survival. We then fit a reduced model to estimate the bias in the logistic regression coefficient of chamber diameter in each food web.

```{r Bias in food-web model, cache=TRUE}
biased_foodweb_df <- gall_selection.df %>%
  group_by(Foodweb, Gall_Number) %>%
  mutate(mean_survival = mean(gall_survival)) %>%
  filter(mean_survival > 0, mean_survival < 1) %>%
  ungroup()

# using linear model, but I'm only interested in the chamber diameter coefficient
biased_foodweb_model <- glmer(
  gall_survival ~ -1 + Foodweb + 
    Foodweb:sc.Diam + 
    (1|Genotype/Plant_Position/Gall_Number),
  data = biased_foodweb_df,
  family = binomial(link = logit), control=glmerControl(optimizer = "bobyqa"))

biased_foodweb_confint <- tidy(biased_foodweb_model, conf.int=TRUE) %>% filter(group=="fixed")

knitr::kable(biased_foodweb_confint)
```

```{r Food-web alpha coefficients, include=FALSE}
foodweb_bind <- bind_cols(
  as_tibble(boot_linear_foodweb_model$t),
  select(as_tibble(boot_foodweb_model$t), `FoodwebComplex:I(sc.Diam^2)`:`FoodwebSimple:sc.log.Clutch:sc.sqrt.Pref`) # only retain nonlinear coefficients from full model
)

@@

```{r Table of food web coefficients, echo=F}
knitr::kable(tidy_foodweb_alphas)
```

```{r Plot food-web coefficients, echo=FALSE}

@@

```{r Food-web selection gradients, include=FALSE}

@@

@@

```{r Plot food-web gradients, echo=FALSE}

@@

egglarval_model <- update(foodweb_model, data=egglarval_df)

biased_egglarval_df <- egglarval_df %>%
  group_by(Foodweb, Gall_Number) %>%
  mutate(mean_survival = mean(gall_survival)) %>%
  filter(mean_survival > 0, mean_survival < 1) %>%
  ungroup()

biased_egglarval_model <- update(biased_foodweb_model, data=biased_egglarval_df)
```

```{r Bootstrap egg vs. larval ptoid model, cache=TRUE, include=FALSE}
boot_egglarval_model <- bootMer(egglarval_model, FUN = fixef, nsim=n_boots_analysis, parallel="multicore", ncpus=32)
```

@@

```{r Bootstrap linear egg vs. larval model, cache=TRUE, include=FALSE}
boot_linear_egglarval_model <- bootMer(linear_egglarval_model, FUN = fixef, nsim=n_boots_analysis, parallel="multicore", ncpus=32)
```

```{r Egg vs. larval ptoid alpha coefficients, include=FALSE}
egglarval_bind <- bind_cols(
  as_tibble(boot_linear_egglarval_model$t),
  select(as_tibble(boot_egglarval_model$t), `FoodwebComplex:I(sc.Diam^2)`:`FoodwebSimple:sc.log.Clutch:sc.sqrt.Pref`) # only retain nonlinear coefficients from full model
)

@@

```{r Tabel of egg vs. larval coefficients, echo=FALSE}
knitr::kable(tidy_egglarval_alphas)
```

```{r Plot egglarval coefficients, echo=FALSE}

@@

```{r Egg vs. larval ptoid selection gradients, include=FALSE}

@@

@@

@@

```{r Figure-Selection-Gradients, echo=FALSE, fig.cap="**Selection gradients**. Estimates of standardized selection gradients in complex (orange) and simple (blue) food webs. The contribution of larval parasitoids (yellow) was estimated with a subset of the complex food-web data that only contained attacks from larval parasitoids (and gall survival). Points and lines correspond to estimates of the mean and 95% confidence intervals, respectively. Overlapping confidence intervals with zero (dotted line) indicate no strong evidence of selection."}
tidy_foodwebegglarval_grads <- bind_rows(
  tidy_foodweb_grads, mutate(filter(tidy_egglarval_grads, type=="Complex"), type="Larval")
)
tidy_foodwebegglarval_grads$type <- factor(tidy_foodwebegglarval_grads$type, levels=c("Complex","Simple","Larval","Diff"))

@@

```{r Plot foodwebegglarval coefficients, echo=FALSE, fig.cap="**Partitioning Selection Gradients.**"}
tidy_foodwebegglarval_alphas <- bind_rows(
  tidy_foodweb_alphas, mutate(filter(tidy_egglarval_alphas, type=="Complex"), type="Larval")
)
tidy_foodwebegglarval_alphas$type <- factor(tidy_foodwebegglarval_alphas$type, levels=c("Complex","Simple","Larval","Diff"))

@@

# Reproduce Figure 4 - Univariate adaptive landscape

```{r Get Gall Diameter Relative Fitness Landscape, cache=TRUE, include=FALSE}
newdata_Diam <- bind_rows(
  expand.grid(Foodweb = "Complex", sc.Diam = seq(-1,1,length.out=1000), sc.log.Clutch = 0, sc.sqrt.Pref = 0),
  expand.grid(Foodweb = "Simple", sc.Diam = seq(-1,1,length.out=1000), sc.log.Clutch = 0, sc.sqrt.Pref = 0))

@@

```{r Get Clutch Size Fitness Landscape, cache=TRUE, include=FALSE}
newdata_Clutch <- bind_rows(
  expand.grid(Foodweb = "Complex", sc.Diam = 0, sc.log.Clutch = seq(-1,1,length.out=1000), sc.sqrt.Pref = 0),
  expand.grid(Foodweb = "Simple", sc.Diam = 0, sc.log.Clutch = seq(-1,1,length.out=1000), sc.sqrt.Pref = 0))

@@

```{r Get Oviposition preference Fitness Landscape, cache=TRUE, include=FALSE}

@@

```{r Plot Univariate Landscapes, include=FALSE}
uni_breaks <- c(0.1,0.5,1.0)

@@

@@

# Get legend
AF_uni_legend <- get_legend(AF_uni_Diam + theme(legend.title.align = 0.5)) #get_legend(AF_uni_Diam)

@@

```{r Figure-Univariate-Landscapes, echo=FALSE, eval=FALSE}
# Get legend
AF_uni_legend <- get_legend(legend_only + theme(legend.position="top", legend.title.align = 0.5) + guides(color=guide_legend(title.position="top")))

@@

```{r Alt Figure-Univariate-Landscapes, echo=FALSE, eval=FALSE}
# Get legend
AF_uni_legend <- get_legend(legend_only + theme(legend.title.align = 0.5))

@@

```{r Egg-Parasitoid-Selection-linear, echo=F}
linear_eggptoid_plot_df <- linear_eggegg_RF_Pref$absolute_fitness %>% 
  select(-sc.Diam, -sc.log.Clutch) %>%
  gather(ID, absolute_fitness, -sc.sqrt.Pref, -Foodweb) %>%
  mutate(ID_group = ifelse(ID == "average", "average", "replicate")) %>%
  spread(Foodweb, absolute_fitness) %>%
  mutate(fitness_difference = Complex-Simple)

@@

# Multivariate fitness landscapes

```{r Get Clutch Size x Oviposition preference Landscape, cache=TRUE, include=FALSE}

@@

```{r Get Gall Diameter x Clutch Size Landscape, cache=TRUE, include=FALSE}

@@

```{r Get Gall Diameter x Oviposition preference Landscape, cache=TRUE, include=FALSE}

@@

```{r Get plots of multivariate adaptive landscape, include=FALSE}

@@

Reproduce Figure 3 in main manuscript.

```{r Figure-Multivariate-Landscapes, echo=FALSE, fig.cap="Fitness landscapes of gall traits in complex vs. simple food webs. Each panel corresponds to a different combination of traits: clutch size and gall diameter (A); clutch size and Oviposition preference (B); Oviposition preference and gall diameter (C). Note that traits for all plots range 1 SD below and above the mean (=0)."}

@@

```{r Slope and Curvature of the Adaptive Landscape, include=FALSE}

@@

>>>>

---
title: "Reproduce analyses reported in: *Phenotypic evolution is more constrained in simpler food webs*"
author: "Matthew A. Barbour"
date: "`r Sys.Date()`"
output: github_document
fontsize: 12pt
---

@@

## Load required libraries ----
library(car)        # for homogeneity of variance test
library(tidyverse)  # for managing data
library(cowplot)    # pretty default ggplots
library(broom)      # for tidying multiple linear models
library(visreg)     # for quick visualizations of model effects
library(viridis)    # for color palette
library(lme4)       # for generalized linear mixed models
library(piecewiseSEM) # for calculating model R-square
library(latex2exp)  # using Latex notation for axes labels
library(stringr)
library(MVN)        # assessing multivariate normality assumption

@@

We write the model in a way that independently estimates the effect of food-web treatment, each trait, and all two-way and three-way statistical interactions, on larval survival. The resulting estimates and confidence intervals are useful for determining whether trait-fitness relationships differ from zero, but not whether they differ between food-web treatments. For the later, we calculate the differences between each food-web treatment from the bootstrapped samples.

@@

We used parametric bootstrapping to calculate 95% confidence intervals of trait-fitness relationships.

```{r Bootstrap food-web model, cache=TRUE}
boot_foodweb_model <- bootMer(foodweb_model, FUN = fixef, nsim=n_boots_analysis, parallel="multicore", ncpus=32)
```

In order to reliably quantify linear trait-fitness relationships, we remove all higher-order terms from the model [Stinchcombe et al. 2008](https://onlinelibrary.wiley.com/doi/full/10.1111/j.1558-5646.2008.00449.x).

@@

And refit using parametric bootstrapping.

```{r Bootstrap linear food-web model, cache=TRUE}
boot_linear_foodweb_model <- bootMer(linear_foodweb_model, FUN = fixef, nsim=n_boots_analysis, parallel="multicore", ncpus=32)
```

Chamber diameter, but not the other phenotypes, could be influenced by parasitism itself rather than being under natural selection. To estimate this potential bias, we subset our data to only include multi-chambered galls where there was variability in larval survival. We then fit a reduced model to estimate the bias in the logistic regression coefficient of chamber diameter in each food web.

```{r Bias in food-web model, cache=TRUE}

# Subset data
biased_foodweb_df <- gall_selection.df %>%
  group_by(Foodweb, Gall_Number) %>%
  mutate(mean_survival = mean(gall_survival)) %>%
  filter(mean_survival > 0, mean_survival < 1) %>%
  ungroup()

# Fit linear model with only chamber diameter 
biased_foodweb_model <- glmer(
  gall_survival ~ -1 + Foodweb + 
    Foodweb:sc.Diam + 
    (1|Genotype/Plant_Position/Gall_Number),
  data = biased_foodweb_df,
  family = binomial(link = logit), control=glmerControl(optimizer = "bobyqa"))

# Accuracy of confidence intervals is not a priority here, so we just use the asymptotic estimates rather than parametric bootstrapping.
biased_foodweb_confint <- tidy(biased_foodweb_model, conf.int=TRUE) %>% filter(group=="fixed")

# Print table
knitr::kable(biased_foodweb_confint)
```

We gather regression coefficients from the full and reduced model to display trait-fitness relationships. For chamber diameter, we subtracted the coefficient due to bias to better approximate the trait-fitness relationship. We also used the bootstrapped samples to calculate the differences in estimates between treatments (labelled **Diff**). The code for tidying this data is not displayed, but is available in the 'reproduce_analyses.Rmd'.

```{r Food-web alpha coefficients}
foodweb_bind <- bind_cols(
  as_tibble(boot_linear_foodweb_model$t),
  select(as_tibble(boot_foodweb_model$t), `FoodwebComplex:I(sc.Diam^2)`:`FoodwebSimple:sc.log.Clutch:sc.sqrt.Pref`) # only retain nonlinear coefficients from full model
)

@@

Table of trait-fitness relationships:

```{r Table of food web coefficients}
knitr::kable(tidy_foodweb_alphas)
```

Plot these trait-fitness relationships:

```{r Plot food-web coefficients}

@@

Using the methods proposed by [Janzen and Stern, 1998](https://onlinelibrary.wiley.com/doi/10.1111/j.1558-5646.1998.tb02237.x), we estimated selection gradients from our trait-fitness relationships.

```{r Food-web selection gradients}

@@

Table of selection gradients.

@@

Plot of selection gradients.

```{r Plot food-web gradients}

@@

# fit model with same structure as foodweb_model
egglarval_model <- update(foodweb_model, data=egglarval_df)

# subset data to estimate bias
biased_egglarval_df <- egglarval_df %>%
  group_by(Foodweb, Gall_Number) %>%
  mutate(mean_survival = mean(gall_survival)) %>%
  filter(mean_survival > 0, mean_survival < 1) %>%
  ungroup()

# fit model with same structure as biased_foodweb_model
biased_egglarval_model <- update(biased_foodweb_model, data=biased_egglarval_df)
```

Parametric bootstrapping to estimate confidence intervals

```{r Bootstrap egg vs. larval ptoid model, cache=TRUE}
boot_egglarval_model <- bootMer(egglarval_model, FUN = fixef, nsim=n_boots_analysis, parallel="multicore", ncpus=32)
```

Fit a reduced model to reliably estimate linear trait-fitness relationships.

@@

Parametric bootstrapping to estimate confidence intervals.

```{r Bootstrap linear egg vs. larval model, cache=TRUE}
boot_linear_egglarval_model <- bootMer(linear_egglarval_model, FUN = fixef, nsim=n_boots_analysis, parallel="multicore", ncpus=32)
```

Tidy trait-fitness relationships for egg and larval parasitoids from above models.

```{r Egg vs. larval ptoid alpha coefficients}
egglarval_bind <- bind_cols(
  as_tibble(boot_linear_egglarval_model$t),
  select(as_tibble(boot_egglarval_model$t), `FoodwebComplex:I(sc.Diam^2)`:`FoodwebSimple:sc.log.Clutch:sc.sqrt.Pref`) # only retain nonlinear coefficients from full model
)

@@

Table of trait-fitness relationships for egg and larval parasitoids.

```{r Tabel of egg vs. larval coefficients}
knitr::kable(tidy_egglarval_alphas)
```

Plot of trait-fitness relationships for egg and larval parasitoids.

```{r Plot egglarval coefficients}

@@

Calculate selection gradients from trait-fitness relationships.

```{r Egg vs. larval ptoid selection gradients}

@@

Table of selection gradients imposed by egg and larval parasitoids.

@@

Plot of selection gradients imposed by egg and larval parasitoids.

@@

```{r Figure-Selection-Gradients, echo=FALSE, fig.cap="**Selection gradients**. Estimates of standardized selection gradients in complex (orange) and simple (blue) food webs. The contribution of larval parasitoids (yellow) was estimated with a subset of the complex food-web data that only contained attacks from larval parasitoids (and gall survival). Points and lines correspond to estimates of the mean and 95% confidence intervals, respectively. Overlapping confidence intervals with zero (dotted line) indicate no strong evidence of selection.", include=FALSE}
tidy_foodwebegglarval_grads <- bind_rows(
  tidy_foodweb_grads, mutate(filter(tidy_egglarval_grads, type=="Complex"), type="Larval")
)
tidy_foodwebegglarval_grads$type <- factor(tidy_foodwebegglarval_grads$type, levels=c("Complex","Simple","Larval","Diff"))

@@

```{r Plot foodwebegglarval coefficients, echo=FALSE, fig.cap="**Partitioning Selection Gradients.**", include=FALSE}
tidy_foodwebegglarval_alphas <- bind_rows(
  tidy_foodweb_alphas, mutate(filter(tidy_egglarval_alphas, type=="Complex"), type="Larval")
)
tidy_foodwebegglarval_alphas$type <- factor(tidy_foodwebegglarval_alphas$type, levels=c("Complex","Simple","Larval","Diff"))

@@

# Reproduce Figure 2 in main text of manuscript.

Use the fitted model to generate predicted estimates of mean fitness for changes in the mean trait value of a population. I restrict these predictions to +/- 1 SD because we can only reliably estimate the shape of the adaptive landscape near the mean phenotype of the population [Arnold et al., 2001](https://link.springer.com/article/10.1023/A:1013373907708). Other trait values are held constant at the mean phenotype (i.e. trait = 0).

```{r Get Gall Diameter Relative Fitness Landscape, cache=TRUE}
newdata_Diam <- bind_rows(
  expand.grid(Foodweb = "Complex", sc.Diam = seq(-1,1,length.out=1000), sc.log.Clutch = 0, sc.sqrt.Pref = 0),
  expand.grid(Foodweb = "Simple", sc.Diam = seq(-1,1,length.out=1000), sc.log.Clutch = 0, sc.sqrt.Pref = 0))

@@

```{r Get Clutch Size Fitness Landscape, cache=TRUE}
newdata_Clutch <- bind_rows(
  expand.grid(Foodweb = "Complex", sc.Diam = 0, sc.log.Clutch = seq(-1,1,length.out=1000), sc.sqrt.Pref = 0),
  expand.grid(Foodweb = "Simple", sc.Diam = 0, sc.log.Clutch = seq(-1,1,length.out=1000), sc.sqrt.Pref = 0))

@@

```{r Get Oviposition preference Fitness Landscape, cache=TRUE}

@@

Create plots for each panel of Figure 2 in main text.

```{r Plot Univariate Landscapes}
uni_breaks <- c(0.1,0.5,1.0)

@@

Fit plots together to reproduce Figure 2.

@@

# Get legend
AF_uni_legend <- get_legend(AF_uni_Diam + theme(legend.title.align = 0.5))

@@

```{r Figure-Univariate-Landscapes, echo=FALSE, eval=FALSE, include=F}
# Get legend
AF_uni_legend <- get_legend(legend_only + theme(legend.position="top", legend.title.align = 0.5) + guides(color=guide_legend(title.position="top")))

@@

```{r Alt Figure-Univariate-Landscapes, echo=FALSE, eval=FALSE, include=F}
# Get legend
AF_uni_legend <- get_legend(legend_only + theme(legend.title.align = 0.5))

@@

```{r Egg-Parasitoid-Selection-linear, echo=F, eval=F, include=F}
linear_eggptoid_plot_df <- linear_eggegg_RF_Pref$absolute_fitness %>% 
  select(-sc.Diam, -sc.log.Clutch) %>%
  gather(ID, absolute_fitness, -sc.sqrt.Pref, -Foodweb) %>%
  mutate(ID_group = ifelse(ID == "average", "average", "replicate")) %>%
  spread(Foodweb, absolute_fitness) %>%
  mutate(fitness_difference = Complex-Simple)

@@

# Reproduce Figure 3 presented in main text.

Use the fitted model to generate predicted estimates of mean fitness for different phenotypic combinations. I restrict these predictions to +/- 1 SD because we can only reliably estimate the shape of the adaptive landscape near the mean phenotype of the population [Arnold et al., 2001](https://link.springer.com/article/10.1023/A:1013373907708).

```{r Get Clutch Size x Oviposition preference Landscape, cache=TRUE}

@@

```{r Get Gall Diameter x Clutch Size Landscape, cache=TRUE}

@@

```{r Get Gall Diameter x Oviposition preference Landscape, cache=TRUE}

@@

Create plots for each panel.

```{r Get plots of multivariate adaptive landscape}

@@

Fit plots together to reproduce Figure 3.

```{r Figure-Multivariate-Landscapes}

@@

# Calculating Slope and Curvature of the Adaptive Landscape

Below is the code I used to get estimates for the slope and curvature of the adaptive landscape in each food-web treatment. If 95% confidence intervals of selection gradients overlap zero, then they are set to zero; otherwise, I retain the mean estimate.

```{r Slope and Curvature of the Adaptive Landscape}

@@

I use these estimates to populate the curvature matrices presented in the **Results** of the main text. I reproduce these estimates here for clarity:

$$\textbf{C} = \begin{pmatrix} C_{\text{Diam:Diam}}&& \\ C_{\text{Clutch:Diam}} & C_{\text{Clutch:Clutch}} & \\ C_{\text{Pref:Diam}} & C_{\text{Pref:Clutch}} & C_{\text{Pref:Pref}} \end{pmatrix}$$

$$\textbf{C}_{\text{Complex}} = \begin{pmatrix} 
`r round(complex_curvature["Diam","Diam"],2)` &  &  \\  
`r round(complex_curvature["Clutch","Diam"],2)` & `r round(complex_curvature["Clutch","Clutch"],2)` &  \\  
`r round(complex_curvature["Pref","Diam"],2)` & `r round(complex_curvature["Pref","Clutch"],2)` & `r round(complex_curvature["Pref","Pref"],2)` \end{pmatrix}$$

$$\textbf{C}_{\text{Simple}} = \begin{pmatrix} 
`r round(simple_curvature["Diam","Diam"],2)` &  &  \\  
`r round(simple_curvature["Clutch","Diam"],2)` & `r round(simple_curvature["Clutch","Clutch"],2)` &  \\  
`r round(simple_curvature["Pref","Diam"],2)` & `r round(simple_curvature["Pref","Clutch"],2)` & `r round(simple_curvature["Pref","Pref"],2)` \end{pmatrix}$$
